ODGI: tools for working with pangenome graphs

### 2021-08-11 manuscript/main.tex
stuff

\title[short Title]{ODGI: scalable tools for working with pangenome graphs}
\author[Heumos, \textit{et~al}.]{
Simon Heumos\,$^{\text{\sfb 1, \sfb 2} \dagger}$,
Andrea Guarracino\,$^{\text{\sfb 2} \dagger}$,
Pjotr Prins\,$^{\text{\sfb 3}}$,
and Erik Garrison\,$^{\text{\sfb 3}*}$
}

@@

\abstract{
\textbf{Motivation:} Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
These models, therefore, offer the opportunity to study the entire genetic diversity of a population, bringing new challenges to how to analyze hundreds of genomes on a gigabase scale.
In addition, vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
Efficient and flexible tools are therefore required to work on pangenome graphs of any complexity and scale. \\
\textbf{Results:} Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
These models, therefore, offer the opportunity to study the entire genetic diversity of a population, bringing new challenges to how to analyze hundreds of genomes on a gigabase scale.
In addition, vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
Efficient and flexible tools are therefore required to work on pangenome graphs of any complexity and scale. \\
\textbf{Availability:} ODGI is written in C++. Its source code is freely available at \url{https://github.com/pangenome/odgi} with documentation at \url{https://odgi.readthedocs.io}.
ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or Guix https://github.com/ekg/guix-genomics/blob/master/odgi.scm. \\
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}

@@

\begin{methods}

@@

ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient. It applies concepts first introduced in the dynamic version of the Graph BWT {31406990} (gBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform {28702075}. Specifically, we build a node-local model of the graph, where the graph is stored in a vector of node structures. The identifier of each node is its offset in this vector plus 1. Each structure contains the sequence of the node, its edges in both directions (forward and backward), and a vector of path steps that describes how the embedded paths walk across the node. Each step records the path identifier and the previous and next path steps on the same path. To reduce the encoding, we encode path steps using a local alphabet that maps the N neighbors of the node into the range 1..N. To further save space, node deltas, rather than IDs, are stored in this alphabet. The delta between two nodes is defined as their distance in the graph vector (i.e., the difference between the node offsets). The path step vector, of length n, is stored using a dynamic succinct integer vector that requires O(5nw) bits, where w is ~log2(N). Finally, to allow operating on path steps in parallel, each node structure includes a byte-width mutex lock.

\end{methods}

@@

\subsection{Core functionality}

@@

\begin{comment}

% the table

Table 1 - A selection of the tools available in ODGI
Name
Description
For conversion

build
Convert graphs from GFA to ODGI format.
view
Convert graphs from ODGI to GFA format.
For validation

validate
Check if the graph is consistent.
For interrogation

stats
Return graph metrics.
bin
Summarize path information in bins of fixed size.
depth
Return node depth as defined by query criteria.
degree
Return node degree as defined by query criteria.
For manipulation

break
Break cycles in the graph, reducing its complexity.
groom
Resolve spurious inverting links.
chop
Chop long nodes into short ones while preserving graph topology and node order.
unchop
Join nodes that can be merged without changing the graph topology, nor the embedded sequences.
explode
Separate the graph into the connected components.
squeeze
Squeeze multiple graphs into the same file, preventing node ID collisions.
extract
Extract parts of the graph as defined by query criteria.
For coordinate projections

position
Coordinate projections between nodes and paths in different graphs.
For visualization

viz
Visualize the graph in one dimension.
sort
Sort the graph.
layout
Generate two-dimension graph layouts.
draw
Draw two-dimension layouts.

\end{comment}

@@

\subsection{Performance on human data}

Table 2 shows a comparison of the operations necessary to extract a subgraph using ODGI and VG. The example consists in extracting the centromeres from a pangenome graph made from 88 haploid, phased human genome assemblies from 44 individuals from the Human Pangenome Reference Consortium year 1 assembly (see Data Availability), plus the chm13 and GRCh38 reference genomes, for a total of 90 haplotypes. All commands are provided in the Supplementary Material.

To work with ODGI and VG it is first necessary to convert the graph from GFA to ODGI and VG format, respectively. Thanks to the ODGI format’s path implementation, odgi build embeds the paths in parallel, outperforming vg convert in the same task. For the same reason, odgi extract results 20 times faster than vg chunk in extracting a portion from the full graph.

\begin{comment}

% table on human perf

Table 2 - Centromere extraction comparison using ODGI and VG
Operation
Tool
Runtime
(mm:ss)
Memory
(GB)
Output size
(GB)
Format conversion

odgi build
1:35
10,39
5,4

vg convert
8:14
28,43
6,1
Subgraph extraction

odgi extract
1:15
9,43
2,7

vg chunk
21:09
59,33
1,7

\end{comment}

@@

>>>>

\title[short Title]{ODGI: tools for working with pangenome graphs}
\author[Heumos, \textit{et~al}.]{
Simon Heumos\,$^{\text{\sfb 1, \sfb 2} \dagger}$,
Andrea Guarracino\,$^{\text{\sfb 2} \dagger}$,
Pjotr Prins\,$^{\text{\sfb 3}}$,
and Erik Garrison\,$^{\text{\sfb 3}*}$
}

@@

% XXX key message of the paper is that we have collected a set of algorithms that enable easy use of pangenome graphs for investigating biology

\abstract{
\textbf{Motivation:} Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
These models offer the opportunity to study the total genomic diversity of a population, including structurally complex regions.
Analyzing hundreds of gigabase-scale genomes using this model is difficult and not well-supported by existing tools.
%In addition, vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
Efficient and flexible tools are therefore required to work on pangenome graphs of any complexity and scale. \\
\textbf{Results:} Here we present ODGI, a new suite of tools to analyze and manipulate pangenome graphs structured in the variation graph model.
ODGI supports graphs in the Graphical Fragment Assembly format version 1 (GFAv1).
The tools scale efficiently to large collections of eukaryotic genomes, allowing users to work easily with pangenome graphs of tens to thousands of large genomes while maintaining fast execution runtime and low memory overhead.
\textbf{Availability:} ODGI is written in C++. Its source code is freely available at \url{https://github.com/pangenome/odgi} with documentation at \url{https://odgi.readthedocs.io}.
ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or Guix https://github.com/ekg/guix-genomics/blob/master/odgi.scm. \\
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}

@@

%\begin{methods}

@@

ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
It applies concepts first introduced in the dynamic version of the Graph BWT {31406990} (gBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform {28702075}.
Specifically, we build a node-local model of the graph, where the graph is stored in a vector of node structures.
The identifier of each node is its offset in this vector plus 1.
Each structure contains the sequence of the node, its edges in both directions (forward and backward), and a vector of path steps that describes how the embedded paths walk across the node.
Each step records the path identifier and the previous and next path steps on the same path.
To reduce the encoding, we encode path steps using a local alphabet that maps the N neighbors of the node into the range $1\ldots N$.
To further save space, node deltas, rather than IDs, are stored in this alphabet.
The delta between two nodes is defined as their distance in the graph vector (i.e., the difference between the node offsets).
The path step vector, of length n, is stored using a dynamic succinct integer vector that requires O(5nw) bits, where w is ~log2(N).
Finally, to allow operating on path steps in parallel, each node structure includes a byte-width mutex lock.

%\end{methods}

@@

%\subsection{Core functionality}

@@

@@

\begin{comment}
key message of the paper is that we have collected a set of algorithms that enable easy use of pangenome graphs for investigating biology
-> build model solves problem of working with big graphs in memory
-> view (convert to GFA) & paths solve problem of exporting basic features of the graph (e.g. paths)
-> stats (understand basic size / structure) & bin & degree & depth solves problem of understanding the overall structure and size of the graph
-> sort (groom) & layout solves problem of finding latent structure in the pangenome
-> viz & draw provides a human-viewable readout of the graph
-> chop & unchop & squeeze & break & prune & explode lets us break apart or combine the graph nodes and topology
-> position & tips & untangle (jaccard based coordinate conversion) provides a way to map coordinates between any genomes in the graph (e.g. liftover!)
-> extract lets us pull out specific regions of the graph based on path ranges, nodes and positions
\end{comment}

\subsection{Building the \textsc{ODGI} model}

% erik

% build
Solves problem of working with big and complex graphs in limited memory.

Figure: encoding model as a cartoon --  HTT exon 1 as tiny example graph (maybe)

\subsection{Converting to standard text formats}

% pjotr

% view / paths
Allows us to export basic data of the graph (genomes and graph structure).

Figure: .dot showing the relationship between all the tools and the data formats we use.

\subsection{Obtaining metrics of the pangenome graph}

% simon
% chr8 longs

% stats / bin / degree / depth / multiQC
Solves the problem of understanding the overall structure, size, and features of the graph and its sequences.

Figure: small graph with all the stats shown. HTT exon 1 as tiny example graph

\subsection{Finding latent structure in the pangenome graph topology}

% simon

% sort / layout / tension??
Allows us to understand the sparse structures typically found in pangenome graphs.

\subsection{Visualizing pangenome graphs}

% andrea (viz) & erik (draw)

% viz / draw
These provide scalable ways of geneating human-viewable pictures of the high-level structure of the graph.
% interesting that we don't go base-level at all... we use vg for that

Figure here with 1D + 2D vizs.

\subsection{Editing the graph structure}

% andrea (?)

% chop / unchop / squeeze / break / prune / explode
These are commonly-needed basic operations on the topology of the graph.

\subsection{Untangling and navigating the pangenome with path-based coordinates}

% pjotr

% position / tips
% untangle (this follows on the jaccard graph mapping concepts in the previous section}
The graph is a model of an alignment of many genomes.
We can use coordinates in any genome to refer to it, but this requires a few basic operations ...
Obtaining unambiguous mappings between different genomes requires the use of a new kind of graph based mapping (path graph jaccard).

Figure showing the untangling of C4 ?
Figure showing tips over a single human chromosome from HPRC.

\subsection{Extracting regions of interest}

% andrea (?)

% extract
Pangenome graphs are very large, but we often only want to work with a small portion (e.g. a single gene).
We can extract such regions using coordinates on the paths in the graph to guide us.

Figure showing extraction process---either schematic or ``real''.

@@

% erik & pjotr & anyone
